Achieving maximum possible precision on stellar radial-velocity measurements in multi-epoch spectroscopy

### 2017-12-02 paper/paper.tex
cosmetics
more on CRLB

\newcommand{\unit}[1]{\mathrm{#1}}
\newcommand{\km}{\unit{km}}
\newcommand{\m}{\unit{m}}
\newcommand{\s}{\unit{s}}
\newcommand{\kms}{\km\,\s^{-1}}
\newcommand{\ms}{\m\,\s^{-1}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\HARPS}{\project{\acronym{HARPS}}}
\newcommand{\RV}{\acronym{RV}}
\newcommand{\CRLB}{\acronym{CRLB}}
\newcommand{\EPRV}{\acronym{EPRV}}
\newcommand{\dd}{\mathrm{d}}

\shorttitle{achieving maximum radial-velocity precision}
\shortauthors{bedell and hogg}

@@

\begin{abstract}
% context
Extremely precise stellar radial velocity (\RV) measurements are a foundational tool in the field of exoplanets, with hundreds of planets discovered by the Doppler method to date and about a dozen high-resolution spectrographs around the world dedicated to the search.
Despite the importance of these observations, the detailed computational methods used to extract $\ms$-level \RV s from stellar spectra remain largely unstandardized and unpublished.
% aims
In this work, we look at sensible approaches to extracting \RV s from a data set of extracted spectra and compare their efficiencies by testing on a segment of realistic model spectrum, with the goal of identifying methods that have the possibility of saturating the information-theoretic Cram\'er--Rao bound (\CRLB) on measurement precision.
% methods
We consider cross-correlations (and, equivalently, maximum-likelihood optimizations) of the data with generalized (weighted) binary masks, synthetic spectra (from, say, stellar models), and data-driven templates constructed from the data themselves.
We also consider the possibility that there are low-amplitude, unknown telluric absorptions (micro-tellurics) affecting the spectrum, and continuum-normalization problems, and measure improvements that flow from modeling those also in a data-driven way.
% results
We find that of all methods, only cross-correlation with an iteratively built data-driven template comes close to saturating the \CRLB.
We find that if there are micro-tellurics, they can be captured by a data-driven model.
We predict that simultaneously fitting data-driven models of the stellar spectrum and telluric contributions at all observational epochs will improve the \RV\ precision of \HARPS\ to substantially better than $1\,\ms$.
\end{abstract}

@@

Given finite, noisy data, there are limits coming from information
theory on how well any parameter of interest can be measured.
The most straightforward bound is the Cram\'er--Rao Lower Bound (\CRLB) on the
variance of any unbiased frequentist estimator.
Estimators can only beat this bound by taking on bias.

This bound is strong, but it cannot be calculated \emph{objectively};
any description of the information in a data set requires
subjective decisions (that is, debatable decisions), which are inputs
to a model for the generation of the data.
In particular, information-theory bounds require a model for the
probability of the data given the parameter of interest; that is, they
require a likelihood function.
(Side note: Bayesians need a likelihood function for inference, and
frequentists need a likelihood function for analyzing their estimators.
The likelihood function is central to all statistical methods.)

In particular, the \CRLB\ is related to the Fisher Information that is
frequently used in cosmological experiment forecasts (\todo{CITE
  SHIT}).
It is that the \emph{inverse variance} $\sigma_v^{-2}$ of any velocity
estimator must be smaller than the second derivative of the
\emph{negative} log likelihood function:
\begin{eqnarray}
\frac{1}{\sigma_v^2} &\geq& \frac{\dd^2}{\dd v^2}(-\ln L)
\end{eqnarray}

In the case of \EPRV, this says that any computation of the limits
on estimating stellar velocities (or stellar velocity changes)
require making assumptions about the stellar spectrum (the
latent, unobserved, \emph{true} spectrum that could only be observed
with arbitrarily good data) and the noise processes that distort the
spectrum in any finite, noisy observation.
These noise processes ought (in principle) to include any beliefs about
unsubtracted telluric absorptions or emissions, and time variability
of the star itself, not to mention photon and spectrograph read and
extraction noise.
However---and in keeping with the literature---we can make maximal or
best-case assumptions, and ask what we would get for \EPRV\ precision
in that best case.
Then we can ask (with fake data) how much different choices about our
data analysis, and different adversarial injections of noise,
tellurics, spectrograph issues, and other wrongness can prevent us
from achieving the best-case bounds.

@@

\begin{itemize}
\item the Cramer-Rao bound is the limit on measurement
\item it relates to Fisher information
\item it cannot be computed without a model
\item CRLB is saturated with a maximum likelihood estimator under the correct model
\item wrong spectrum template = wrong and noisy \RV s
\end{itemize}

\begin{itemize}
\item what is the Cramer-Rao bound on a spectrum? See notes in Hogg's red book for 2017-12-01.
\item Butler equation
\item Bouchy equation
\item Figueira (or Lovis \& Fischer) equation
\end{itemize}

>>>>

% units macros
\newcommand{\unit}[1]{\mathrm{#1}}
\newcommand{\km}{\unit{km}}
\newcommand{\m}{\unit{m}}
\newcommand{\s}{\unit{s}}
\newcommand{\kms}{\km\,\s^{-1}}
\newcommand{\ms}{\m\,\s^{-1}}

% math macros
\newcommand{\dd}{\mathrm{d}}
\newcommand{\T}{^{\mathsf{T}}}

% text macros
\newcommand{\todo}[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\HARPS}{\project{\acronym{HARPS}}}
\newcommand{\RV}{\acronym{RV}}
\newcommand{\CRLB}{\acronym{CRLB}}
\newcommand{\EPRV}{\acronym{EPRV}}

\setlength{\parindent}{1.4em}
\shorttitle{achieving maximum radial-velocity precision}
\shortauthors{bedell and hogg}

@@

\begin{abstract}\noindent
% context
Extremely precise stellar radial velocity (\RV) measurements are a foundational tool in the field of exoplanets, with hundreds of planets discovered by the Doppler method to date and about a dozen high-resolution spectrographs around the world dedicated to the search.
Despite the importance of these observations, the detailed computational methods used to extract $\ms$-level \RV s from stellar spectra remain largely unstandardized and unpublished.
% aims
In this work, we look at sensible approaches to extracting \RV s from a data set of extracted spectra and compare their efficiencies by testing on a segment of realistic model spectrum, with the goal of identifying methods that have the possibility of saturating the information-theoretic Cram\'er--Rao bound (\CRLB) on measurement precision.
% methods
We consider cross-correlations (and, equivalently, maximum-likelihood optimizations) of the data with generalized (weighted) binary masks, synthetic spectra (from, say, stellar models), and data-driven templates constructed from the data themselves.
We also consider the possibility that there are low-amplitude, unknown telluric absorptions (micro-tellurics) affecting the spectrum, and continuum-normalization problems, and measure improvements that flow from modeling those also in a data-driven way.
% results
We find that of all methods, only cross-correlation with an iteratively built data-driven template comes close to saturating the \CRLB.
We find that if there are micro-tellurics, they can be captured by a data-driven model.
We predict that simultaneously fitting data-driven models of the stellar spectrum and telluric contributions at all observational epochs will improve the \RV\ precision of \HARPS\ to substantially better than $1\,\ms$.
\end{abstract}

@@

Given finite, noisy data, there is a limit coming from information
theory on how well any parameter of interest can be measured.
This limit is
sometimes known as the Cram\'er--Rao Lower Bound (\CRLB; \todo{cite CR and wikipedia}).
In its simplest form, it constrains the variance of any unbiased
frequentist estimator.
Estimators can only beat this bound by taking on bias.

This bound is strong, but it cannot be calculated \emph{objectively};
any description of the information in a data set requires
subjective decisions (that is, debatable decisions), which are inputs
to a model for the generation of the data.
In particular, information-theory bounds require a model for the
probability of the data given the parameter of interest; that is, they
require a likelihood function.
As an aside: Bayesians need a likelihood function for inference, and
frequentists need a likelihood function for analyzing their estimators.
The likelihood function is therefore central to all statistical methods,
and the core component of any data analysis.

In particular, the \CRLB\ is related to the Fisher Information that is
frequently used in cosmological experiment forecasts (\todo{CITE
  SHIT}).
It is that the \emph{inverse variance} $\sigma_v^{-2}$ of any velocity
estimator (that is, the information in the velocity) must be smaller than the
second derivative of the \emph{negative} log likelihood function:
\begin{equation}
\frac{1}{\sigma_v^2} \leq E\left[\frac{\dd^2}{\dd v^2}(-\ln L)\right] \quad,
\end{equation}
where $L$ is the likelihood, or the probability of the data given the velocity,
and the expectation $E[]$ is taken over all possible data (that is, it is an
expectation under the likelihood).
If there are many other nuisance parameters, a Bayesian can marginalize them
out to make $L$ the marginalized likelihood for the velocity, or a frequentist
can include them in the estimation and compute an inverse covariance matrix $C^{-1}$
for the full parameter vector, which will have the bound
\begin{equation}
v^T\cdot C^{-1}\cdot v \leq v^T\cdot Q\cdot v
\end{equation}
\begin{equation}
Q_{ij} \equiv E\left[\frac{\dd^2}{\dd\theta_i\,\dd\theta_j}(-\ln L)\right] \quad,
\end{equation}
where $Q$ is the information tensor, $i,j$ are indices into that
tensor, corresponding parameters are $\theta_i, \theta_j$, and $v$ is
any arbitrary vector.  That is, the bound is not just on every
parameter, but really on on any combination of parameters.

In the case of \EPRV, this says that any computation of the limits
on estimating stellar velocities (or stellar velocity changes)
requires a likelihood function.
This, in turn, requires making assumptions about the stellar spectrum (the
latent, unobserved, \emph{true} spectrum that could only be observed
with arbitrarily good data) and the noise processes that distort the
spectrum in any finite, noisy observation.
These noise processes ought (in principle) to include any beliefs about
unsubtracted telluric absorptions or emissions, and time variability
of the star itself, not to mention photon and spectrograph read and
extraction noise.
However---and in keeping with the literature---we can make maximal or
best-case assumptions, and ask what we would get for \EPRV\ precision
in that best case.
Then we can ask (with fake data) how much different choices about our
data analysis, and different adversarial injections of noise,
tellurics, spectrograph issues, and other wrongness can prevent us
from achieving the best-case bounds.

@@

\begin{itemize}
\item what is the Cramer-Rao bound on a spectrum in optimistic case? 
\item CRLB is saturated with a maximum likelihood estimator under the correct model
\item wrong spectrum template = wrong and noisy \RV s
\item Butler equation \todo{See notes in Hogg's red book for 2017-12-01.}
\item Bouchy equation
\item Figueira (or Lovis \& Fischer)
\end{itemize}
